---
title: "Expression pattern analysis by gene"
author: "Priyanka Surana"
date: "17/09/2019"
#output:
#  word_document: default
#  html_document:
#    df_print: paged
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Dropbox (Sydney Uni)/RogerWise/PatternAnalysis/")
```

1. Normalize Read Counts
```{r Normalize, echo = FALSE}
# normalized read counts with NA substituted with -0.1
rc = read.csv("geneCounts.csv", row.names = 1)
rc[,91] = NULL
colnames(rc) = gsub("mla6_bln1", "dm", colnames(rc))
rc = rc[c(grep("EPl|AGP|HOR", rownames(rc), value = TRUE)),]
rc = rc[,c(grep("mla6|WT", colnames(rc), value = TRUE))]
rc[] = lapply(rc, as.numeric)
rc = log2(rc)
rc[rc == -Inf] = -0.9
rc[is.na(rc)] = -0.1
```

2. Calculate Averages and Standard Error
```{r AvgsStderrs, echo = FALSE}
# average read counts & standard error
sem = function(x){
	sd(x)/sqrt(length(x))
}

avg_se = function(df){
	n = ncol(df)/3
	df2 = data.frame(row.names = rownames(df))
	df3 = data.frame(row.names = rownames(df))
	for(i in 1:n){
		m = (i-1)*3
		df2[,i] = rowMeans(df[,(m+1):(m+3)])
		df3[,i] = apply(df[,(m+1):(m+3)], 1, sem)
	}
	
	colnames(df2) = paste0(unique(sapply(colnames(df), function(x){ strsplit(x, "_r")[[1]][1] }, USE.NAMES = FALSE)), "_avg")
	df2 = df2[,sort(colnames(df2))]
	
	colnames(df3) = paste0(unique(sapply(colnames(df), function(x){ strsplit(x, "_r")[[1]][1] }, USE.NAMES = FALSE)), "_sem")
	df3 = df3[,sort(colnames(df3))]
	
	df4 = merge(df2, df3, by="row.names")
	colnames(df4)[1] = "gene"
	return(df4)
}

rc2 = avg_se(rc)
```

3. Include Annotation, Differential Expression and Log2 Fold Change Data
```{r DiffExpL2FC, echo = FALSE}
# annotation
ann = read.csv("barleyAnnotation.csv")

# differential expression results
de = read.csv("barleyDiffExp.csv")
de = de[,c("gene", grep("wt_mla6", colnames(de), value = TRUE))]

# log2 fold change results
l2fc = read.csv("barleyL2FC.csv")
l2fc = l2fc[,c("gene", grep("wt_mla6", colnames(l2fc), value = TRUE))]

# final data set
df = merge(rc2, de)
df = merge(df, l2fc)
```

4. Remove High Standard Error Genes
```{r RemoveErrorGenes, echo = FALSE}
# remove genes with high std err.
df2 = df
n = nrow(df2)
gois = c()
for(i in 1:n){
	sems = df2[i,14:25]
	if(length(sems[sems > 1]) == 0){
		gois = c(gois, df2[i,1])
	}
}
df = subset(df2, df2$gene %in% gois)
```

5. Pattern Analysis
The values in each recipe below is set based on particular gene expression profile.

i. ARF2 (HORVU5Hr1G111500)
```{r PatternAnalysisARF2, echo = FALSE}
#subset(df, df$gene == "HORVU5Hr1G111500") #ARF2
#Notes: p1_1 = 26, p1_2 = 1, p1_3 = 567, p1_4 = 85

cat("Method 4: Using mostly log2 fold change (restricted)\n")
cat("Adjusted p-value: <= 0.001 for timepoint 20; No limits set for other timepoints\n")
cat("Log2 fold change: < .5 and > -.5 for timepoints 0, 16, 24, 32, 48; > 1 for timepoint 20\n")
p1_4 = subset(df, df$t0_wt_mla6_log2FoldChange < .5 & df$t0_wt_mla6_log2FoldChange > -.5 & df$t16_wt_mla6_log2FoldChange < .5 & df$t16_wt_mla6_log2FoldChange > -.5 & df$t20_wt_mla6_padj <= 0.001 & df$t20_wt_mla6_log2FoldChange > 1 & df$t24_wt_mla6_log2FoldChange < .5 & df$t24_wt_mla6_log2FoldChange > -.5 & df$t32_wt_mla6_log2FoldChange < .5 & df$t32_wt_mla6_log2FoldChange > -.5 & df$t48_wt_mla6_log2FoldChange < .5 & df$t48_wt_mla6_log2FoldChange > -.5)
cat("Number of genes:", nrow(p1_4), "\n")
cat("Gene names:", p1_4$gene)
```

##### Checking to see if any of the genes from method #4 (Using mostly log2 fold change - restricted) align with Barley1 probesets in 1HS or 2HL eQTL
```{r eqtlARF2, echo = FALSE}
m4 = p1_4$gene
map = read.csv("barleyHarvestMap.csv")
map = subset(map, map$gc_percent_coverage >= 95 & map$gc_percent_identity >= 95)[,c("hv_gene", "probeset")]
eqtl=read.csv("eQTL.csv", header=TRUE)[,1:6]
colnames(eqtl) = c("probeset", "qvalue_16h", "most_associated_marker_16h", "qvalue_32h", "most_associated_marker_32h", "bb4_expression_pattern")
eqtl16 = subset(eqtl, eqtl$qvalue_16h <= 1e-3 & eqtl$most_associated_marker_16h %in% c("2H.63", "2H.65", "2H.66", "2H.67", "2H.68"))
eqtl32 = subset(eqtl, eqtl$qvalue_32h <= 1e-3 & eqtl$most_associated_marker_32h %in% c("1H.03", "1H.04", "1H.05", "1H.06", "1H.07", "1H.08"))
gois = subset(map, map$probeset %in% eqtl16$probeset | map$probeset %in% eqtl32$probeset)
output = subset(gois, gois$hv_gene %in% m4)
output = merge(output, eqtl, by="probeset")
rownames(output) = NULL

cat("Number of overlapping genes:", length(unique(output$hv_gene)), "\n")
cat("Overlapping gene - probeset pairs:", "\n")
output
```

When a gene did not have a significantly different expression profile between *mla6* and wildtype genotypes, the number of similar genes was very high. To reduce this number, genes with low read counts (<= 8 in log2 scale) were removed. Then, the above pattern analysis was repeated.

6. Remove Low Read Count Genes
```{r RemoveLowGenes, echo = FALSE}
df3 = subset(df, df$mla6_0_avg > 8 & df$mla6_16_avg > 8 & df$mla6_20_avg > 8 & df$mla6_24_avg > 8 & df$mla6_32_avg > 8 & df$mla6_48_avg > 8 & df$WT_0_avg > 8 & df$WT_16_avg > 8 & df$WT_20_avg > 8 & df$WT_24_avg > 8 & df$WT_32_avg > 8 & df$WT_48_avg > 8)
cat("Genes in original dataset:", nrow(df2), "\n")
cat("Genes after removing for high standard error:", nrow(df), "\n")
cat("Genes after also removing for low read counts:", nrow(df3), "\n")
df = df3
```

7. Repeat Pattern Analysis
Repeating pattern analysis without low count reads.

i. ARF2 (HORVU5Hr1G111500)
```{r PatternAnalysis2ARF2, echo = FALSE}
#subset(df, df$gene == "HORVU5Hr1G111500") #ARF2
#Notes: p1_1 = 22, p1_2 = 1, p1_3 = 460, p1_4 = 71

cat("Method 4: Using mostly log2 fold change (restricted)\n")
cat("Adjusted p-value: <= 0.001 for timepoint 20; No limits set for other timepoints\n")
cat("Log2 fold change: < .5 and > -.5 for timepoints 0, 16, 24, 32, 48; > 1 for timepoint 20\n")
p1_4 = subset(df, df$t0_wt_mla6_log2FoldChange < .5 & df$t0_wt_mla6_log2FoldChange > -.5 & df$t16_wt_mla6_log2FoldChange < .5 & df$t16_wt_mla6_log2FoldChange > -.5 & df$t20_wt_mla6_padj <= 0.001 & df$t20_wt_mla6_log2FoldChange > 1 & df$t24_wt_mla6_log2FoldChange < .5 & df$t24_wt_mla6_log2FoldChange > -.5 & df$t32_wt_mla6_log2FoldChange < .5 & df$t32_wt_mla6_log2FoldChange > -.5 & df$t48_wt_mla6_log2FoldChange < .5 & df$t48_wt_mla6_log2FoldChange > -.5)
cat("Number of genes:", nrow(p1_4), "\n")
cat("Gene names:", p1_4$gene)
```

##### Checking to see if any of the genes from method #4 (Using mostly log2 fold change - restricted) align with Barley1 probesets in 1HS or 2HL eQTL
```{r eqtl2ARF2, echo = FALSE}
m4 = p1_4$gene
output = subset(gois, gois$hv_gene %in% m4)
output = merge(output, eqtl, by="probeset")
rownames(output) = NULL

cat("Number of overlapping genes:", length(unique(output$hv_gene)), "\n")
cat("Overlapping gene - probeset pairs:", "\n")
output
```
